Client meeting notes
-Theory: Among teammates born in the same year, the oldest (earliest-born) player tends to be physically bigger than peers, which boosts confidence and leads to genuinely better performance
-Client's ask: Test this theory statistically 
-Subject: Canadian hockey players born within the same year
-Outcome variable: Whether the player made it into the NHL / whether the pattern differs by year
-Main predictor: Birth month
-What to look at: Distribution of how many players were born in each month
-Potential confound: The general population itself has more births in January / early in the year, so this could create a false pattern unrelated to the theory — need to account for this baseline
-Scope: Canadian players only
-Data: Two Excel tables sharing player_id, but not the exact same set of players
-Additional hypothesis: Client thinks birth month might relate to height
-Additional question: Whether birth province matters
-Note: Making the NHL itself is already a very high bar  



## Stage 2: Make an analysis plan
- [ ] What model/test/technique you plan to use.

$$\chi^2 = \sum \frac{(observed - expected)^2}{expected}$$


- [ ] A justification for that choice. This should consist of: 
(a) the probabilistic model you think describes the data
:Assumed a Multinomial distribution.
 1. NHL Birth-Month Distribution vs Canadian population(average bm Distribution)
 2. year-matched birth-month distribution -> conclusion consistent   

(b) any simplifications of the model you are imposing to do your test/estimate parameters
:full-period(1991–2022) average birth-month data 

(c) justifications of the simplifications.
:Assumed birth-month pattern stable over time



#Stage3

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```



#- [ ] Get the data on your computer.
```{r}
data_cbirth <- read.csv(r"(C:\Users/hidy0/Desktop/consultation/Roster/canada_births_1991_2022.csv)")
data_cbirth
data_pbirth <- read.csv(r"(C:/Users/hidy0/Desktop/consultation/Roster/nhl_player_births.csv)")
data_pbirth

```



#- [ ] Do quality checks (make sure it looks like what the client said it should).
- Filter for Canadian players only 
-Check for missing values 
```{r}
data_pbirth_can <- data_pbirth[data_pbirth$birth_country == "CAN",]
nrow(data_pbirth_can)

sum(is.na(data_pbirth_can$birth_year))
sum(is.na(data_pbirth_can$birth_month))

sum(is.na(data_cbirth$year))
sum(is.na(data_cbirth$month))
sum(is.na(data_cbirth$births))

```



Don't cover the same time period
-Overlapping period: 91-2005 
```{r}
#full dataset 
table(data_cbirth$year)
table(data_cbirth$month)

table(data_pbirth_can$birth_year)
table(data_pbirth_can$birth_month)
```



Visual comparison
```{r}
prop.table(tapply(data_cbirth$births, data_cbirth$month, sum))
prop.table(table(data_pbirth_can$birth_month))
```
 

 
#- [ ] Get the data into the correct format for the function(s) you are going to use to fit the model.

Analysis1: using the average birth-month distribution

- Expected = total number of players × *the overall proportion*
- Observed = the actual number of NHL players for each month (across all years)
```{r}
observed <- table(data_pbirth_can$birth_month)
expected <- prop.table(tapply(data_cbirth$births, data_cbirth$month, sum))

chisq.test(observed, p=expected)
```

H0: NHL bm distribution = Canada bm distribution





Check: monthly birth proportions are stable across years
```{r}
cbirth_prop <- data_cbirth %>%
  group_by(year) %>%
  mutate(prop = births / sum(births))

ggplot(cbirth_prop,
       aes(year, prop, group=month, color = factor(month))) +
       geom_line() +
         labs(x = "year", y = "Proportion", color = "month")

year_month_table <- xtabs(births ~ year + month, data_cbirth)
chisq.test(year_month_table)
```

-> Canadian birth-month pattern was not exactly the same every year
(= the proportion of babies born each month in Canada can vary from year to year) 


 
---------------------------------------------------------- 



Analysis2: overlapping years(91-2005) -> using year-matched birth distributions
- Expected = that year's total number of players × *that year's Canadian birth proportion*
- Observed = the actual number of NHL players for each year and month.
```{r}
#91-2005 Candanian
exact_cbirth <- data_cbirth[data_cbirth$year >= 1991 
                            & data_cbirth$year <= 2005, ]

#Canadian bm proportions 
exact_cbirth_prop <- exact_cbirth %>%
  group_by(year) %>%
  mutate(prop = births / sum(births))

#91-2005 Candanian players
exact_pbirth <- data_pbirth_can[data_pbirth_can$birth_year>= 1991 & data_pbirth_can$birth_year <= 2005, ]

#Canadian player monthly birth proportions 
exact_pbirth_count <- exact_pbirth %>%
  group_by(birth_year, birth_month) %>%
  summarise(observed = n(), .groups = "drop") %>%
  complete(birth_year = 1991:2005, birth_month = 1:12, fill = list(observed = 0))

exact_pbirth_merge <- exact_pbirth_count %>%
  left_join(exact_cbirth_prop, by = c("birth_year"="year",
                                      "birth_month"="month"))

head(exact_pbirth_merge)
```

Intermediate table: NHL counts & Canadian birth proportions 




```{r}
# number of players per year 
year_totals <- tapply(exact_pbirth_merge$observed, exact_pbirth_merge$birth_year, sum)

exact_pbirth_merge$n_i <- year_totals[as.character(exact_pbirth_merge$birth_year)]

#expected count 
exact_pbirth_merge$expected_ij <- exact_pbirth_merge$n_i * exact_pbirth_merge$prop


#chi
observed_by_month <- tapply(exact_pbirth_merge$observed, exact_pbirth_merge$birth_month, sum)
expected_by_month <- tapply(exact_pbirth_merge$expected_ij, exact_pbirth_merge$birth_month, sum)

chisq.test(observed_by_month, p = expected_by_month / sum(expected_by_month))
```
H0: NHL distribution = Canada distribution

Both analyses -> NHL distribution ≠ Canadian distribution (theory) 






First method (average proportion)

*Residual = Observed−Expected 
```{r}
observed <- table(data_pbirth_can$birth_month)
expected_p <- prop.table(tapply(data_cbirth$births, data_cbirth$month, sum))
result <- chisq.test(observed, p = expected_p)

sort( result$stdres, decreasing=TRUE)
```
Players born earlier in the year are more common in the NHL than expected.





```{r}
stdres_table <- data.frame(month = as.numeric(names(result$stdres)),
                           stdres = as.numeric(result$stdres))

ggplot(stdres_table, aes(x=factor(month), y = stdres)) +
  geom_col() 
```




Check: downward trend  

```{r}
observed_by_month <- as.numeric(table(data_pbirth_can$birth_month))
expected_p <- as.numeric(prop.table(tapply(data_cbirth$births, data_cbirth$month, sum)))
n_total <- sum(observed_by_month)

prop.trend.test(observed_by_month, rep(n_total, 12), score = 1:12)
```
H0: no consistent trend in the monthly proportion
 
 
Result: NHL players were more likely to be born earlier in the year, and less likely to born later in the year. 
-> theory 

 
 
 
 
 
 
 
 
 
 
 
 


## Stage 2: Make an analysis plan

  
- [ ] What model/test/technique you plan to use.
linear regression 

 
## Stage 3: Perform the analysis

- [ ] Get the data on your computer.
```{r}
data_roster <- read.csv(r"(C:/Users/hidy0/Desktop/consultation/Roster/nhl_rosters.csv)")
data_roster
```

- [ ] Do quality checks 
Duplicates in the roster-> removed 
```{r}
data_roster_canada <- data_roster[data_roster$birth_country == "CAN",]

data_roster_can <- data_roster_canada[!duplicated(data_roster_canada$player_id),]

nrow(data_pbirth_can)
nrow(data_roster_can)
```
 

height & birth date 
```{r}
sum(is.na(data_roster_can$height_in_centimeters))
sum(is.na(data_roster_can$birth_date))
```

```{r}
data_roster_can <- data_roster_can[!is.na(data_roster_can$height_in_centimeters), ]

nrow(data_roster_can)
```

 
 
- [ ] Get the data into the correct format for the function(s) you are going to use to fit the model.

:birth_date format is inconsistent -> 160 NA 
```{r}
data_roster_can$birth_month <- as.numeric(format(as.Date(data_roster_can$birth_date,
format = "%m/%d/%Y"), "%m"))

#Birth month gets split into quarters 
data_roster_can$birth_quarter <- cut(data_roster_can$birth_month,
    breaks = c(0, 3, 6, 9, 12),
    labels = c("Q1", "Q2", "Q3", "Q4"))

height_quarter_data <- data_roster_can[, c("height_in_centimeters", "birth_quarter")]

height_quarter_data 
colSums(is.na(height_quarter_data))
```

```{r}
height_quarter_data <- height_quarter_data[!is.na(height_quarter_data$birth_quarter), ]
```




- [ ] Fit the model, get the number(s) you need from the output.
```{r}
model_h <- lm(height_in_centimeters ~ birth_quarter, height_quarter_data)
summary(model_h)
```
- Q4 is 0.57 shorter than Q1: (statistically significant)
- R² = 0.2% 

= Birth quarter explains very little of the height difference




Check regression assumptions
```{r}
plot(model_h)
```





 
 